Climate-driven phytoplankton community variability in the Cariaco Basin, Venezuela

### 2025-06-27 agujournaltemplate.tex
Revised Abstract - post Ago 1st Review

\begin{keypoints}
\item The CARIACO phytoplankton community time series reveals two clusters associated to high- and low-upwelling conditions and corresponding to anomalies in the AMO index. 
\index No significant differences are observed in diversity metrics between the two clusters, with the exception of a lower turnover and a higher community evenness during low-upwelling conditions.
\item Gradient forest analysis identifies the AMO climate index as the strongest predictor of community change, highlighting the influence of large-scale climatic oscillations on this tropical coastal ecosystem.
\end{keypoints}

@@

\begin{abstract}
Phytoplankton communities play a crucial role in oceans' biogeochemical cycles. Climate drives changes in the physical environment that phytoplankton communities experience and can impact, thus, on the productivity and diversity of marine ecosystems. Long-term time series can capture changes in climate and ecosystems, allowing us to untangling of effects on local communities. 
This study aims to describe and investigate the drivers of changes in  phytoplankton community data (biomass and diversity) collected in the Cariaco Basin, Venezuela, between 1995 and 2017. The monthly cruises sampled water for microscopy cell counts and determination of biogeochemical properties, which we integrated with data from AMO and MEI v.2 climate indices and ERA5 climate model reanalysis data. 
Using clustering algorithms, we investigate the structure of the community data, revealing two temporal clusters associated to high and low upwelling conditions (from 1995 to 2004 and from 2005 to 2017, respectively) and showing a return to community biomass and physical conditions similar to those characterizing the period 1995-2004. However,  community diversity did not follow the same pattern, showing a decrease between 1998 and 2009. To quantify the effect of climate and bottom-up drivers on phytoplankton community changes, we employed a gradient forest analysis and found that a combination of in situ, short-term, upwelling-related variables (SST and nitrate concentration) and long-term, climate indices (AMO and MEI v.2) were the strongest predictors. Our study emphasizes the impact of long-term (decadal) climatic oscillations, such as the AMO, on the phytoplankton community of a tropical coastal ecosystem through modulation of the upwelling system and highlights the capability of the ecosystem to recover from climate change.

>>>>

\begin{keypoints}
\item The CARIACO phytoplankton community time series reveals two clusters, separated by a shift from high- to low-upwelling conditions after 2004 and a return to high-upwelling conditions following 2013.
\index No significant differences are observed in diversity metrics between the two clusters, with the exception of a lower turnover and a higher community evenness during low-upwelling conditions.
\item Gradient forest analysis identifies the AMO climate index as the strongest predictor of community change, highlighting the influence of large-scale climatic oscillations on this tropical coastal ecosystem.
\end{keypoints}

@@

\begin{abstract}
Phytoplankton communities play a crucial role in oceans' biogeochemical cycles. Climate drives changes in the physical environment that phytoplankton communities experience, thus impacting the productivity and diversity of marine ecosystems. Long-term time series can capture changes in climate and ecosystems, allowing us to untangle the effects on local communities. 
This study aims to describe and investigate the changes in phytoplankton community in the Cariaco Basin, Venezuela, between 1995 and 2017. Water samples were taken during monthly cruises to determine microscopy cell counts and biogeochemical properties, which we integrated with data from the AMO and MEI v.2 climate indices and ERA5 climate model reanalysis data. 
Using clustering algorithms, we investigate the structure of the community data, revealing two clusters that describe a shift from high- to low-upwelling conditions in 2004, but also show a return to a community and conditions similar to those before the shift in 2014.
However,  community diversity did not follow the same pattern, showing a decrease between 1998 and 2009. To quantify the effect of climate and bottom-up drivers on phytoplankton community changes, we employed a gradient forest analysis and found that a combination of in situ upwelling-related variables (SST and nitrate concentration) and climate indices (AMO and MEI v.2) were the strongest predictors. Our study emphasizes the impact of long-term (decadal) climatic oscillations, such as the AMO, on the phytoplankton community of a tropical coastal ecosystem through modulation of the upwelling system and highlights the capability of ecosystems to recover from the effects of climate change.
